---
title: "Statistical Modelling of COVID-19 Outbreak in Italy"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
    css: "COVID-19-IT.css"
---

<br><br>


```{r setup, echo=FALSE}
source("setup.R")
```

# Nonlinear growth models

Nonlinear growth models represent an instance of *nonlinear regression models*, a class of models taking the general form 
$$
y = \mu(x, \theta) + \epsilon,
$$
where $\mu(x, \theta)$ is the mean function which depends on a possibly vector-valued parameter $\theta$, and a possibly vector-valued predictor $x$.
The stochastic component $\epsilon$ represents the error with mean zero and constant variance. Usually, a Gaussian distribution is also assumed for the error term. 

By defining the mean function $\mu(x, \theta)$ we may obtain several different models, all characterized by the fact that parameters $\theta$ enter in a nonlinear way into the equation. Parameters are usually estimated by *nonlinear least squares* which aims at minimizing the residual sum of squares.


## Exponential

$$
\mu(x) = \theta_1 \exp\{\theta_2 x\}
$$
where $\theta_1$ is the value at the origin (i.e. $\mu(x=0)$), and $\theta_2$  represents the (constant) relative ratio of change (i.e. $\frac{d\mu(x)}{dx }\frac{1}{\mu(x)} = \theta_2$). Thus, the model describes an increasing (exponential growth if $\theta_2 > 0$) or decreasing (exponential decay if $\theta_2 < 0$) trend with constant relative rate.

## Logistic

$$
\mu(x) = \frac{\theta_1}{1+\exp\{(\theta_2 - x)/\theta_3\}}
$$
where $\theta_1$ is the upper horizontal asymptote, $\theta_2$ represents the x-value at the inflection point of the symmetric growth curve, and $\theta_3$ represents a scale parameter (and $1/\theta_3$ is the growth-rate parameter that controls how quickly the curve approaches the upper asymptote).

## Gompertz 

$$
\mu(x) = \theta_1 \exp\{-\theta_2 \theta_3^x\}
$$
where $\theta_1$ is the horizontal asymptote, $\theta_2$ represents the value of the function at $x = 0$ (displacement along the x-axis), and $\theta_3$ represents a scale parameter.

The difference between the logistic and Gompertz functions is that the latter is not symmetric around the inflection point. 

## Richards 

$$
\mu(x) = \theta_1 (1 - \exp\{-\theta_2 x\})^{\theta_3}
$$
where $\theta_1$ is the horizontal asymptote, $\theta_2$ represents the rate of growth, and $\theta_3$  in part determines the point of inflection on the y-axis.


# Data 

Dipartimento della Protezione Civile: COVID-19 Italia - Monitoraggio della situazione <http://arcg.is/C1unv>

Source: https://github.com/pcm-dpc/COVID-19 

```{R}
url = "https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-andamento-nazionale/dpc-covid19-ita-andamento-nazionale.csv"
COVID19 <- read.csv(file = url, stringsAsFactors = FALSE)
COVID19$data <- as.Date(COVID19$data)
DT::datatable(COVID19)
```

Warnings
```
- 29/03/2020: dati Regione Emilia Romagna parziali (dato tampone non aggiornato).
- 26/03/2020: dati Regione Piemonte parziali (-50 deceduti - comunicazione tardiva)
- 18/03/2020: dati Regione Campania non pervenuti.
- 18/03/2020: dati Provincia di Parma non pervenuti.
- 17/03/2020: dati Provincia di Rimini non aggiornati
- 16/03/2020: dati P.A. Trento e Puglia non pervenuti.
- 11/03/2020: dati Regione Abruzzo non pervenuti.
- 10/03/2020: dati Regione Lombardia parziali.
- 07/03/2020: dati Brescia +300 esiti positivi
```

<br>

# Modelling total infected 

```{R}
# create data for analysis
data = data.frame(date = COVID19$data,
                  y = COVID19$totale_casi)
data$x = as.numeric(data$date) - min(as.numeric(data$date)) + 1
DT::datatable(data, options = list("pageLength" = 5))
```

## Estimation

### Exponential

```{R}
mod1_start = lm(log(y) ~ x, data = data)
b = unname(coef(mod1_start))
start = list(th1 = exp(b[1]), th2 = b[2])
exponential <- function(x, th1, th2) th1 * exp(th2 * x)
mod1 = nls(y ~ exponential(x, th1, th2), data = data, start = start)
summary(mod1)
```

### Logistic

```{R}
mod2 = nls(y ~ SSlogis(x, Asym, xmid, scal), data = data)
summary(mod2)
```

### Gompertz 

```{R}
mod3 = nls(y ~ SSgompertz(x, Asym, b2, b3), data = data)
# start = list(Asym = coef(mod2)[1])
# tmp = list(y = log(log(start$Asym) - log(data$y)), x = data$x)
# b = unname(coef(lm(y ~ x, data = tmp)))
# start = c(start, c(b2 = exp(b[1]), b3 = exp(b[2])))
# mod3 = nls(y ~ SSgompertz(x, Asym, b2, b3), data = data, start = start,
#            control = nls.control(maxiter = 1000))
summary(mod3)
```

### Richards

```{R}
richards <- function(x, th1, th2, th3) th1*(1 - exp(-th2*x))^th3
Loss  <- function(th, y, x) sum((y - richards(x, th[1], th[2], th[3]))^2) 
start <- optim(par = c(coef(mod2)[1], 0.001, 1), fn = Loss, 
               y = data$y, x = data$x)$par
names(start) <- c("th1", "th2", "th3")
mod4 = nls(y ~ richards(x, th1, th2, th3), data = data, start = start,
           # trace = TRUE, algorithm = "plinear", 
           control = nls.control(maxiter = 1000, tol = 0.1))
# algorithm is not converging... 
summary(mod4)
# library(nlmrt)
# mod4 = nlxb(y ~ th1*(1 - exp(-th2*x))^th3, 
#             data = data, start = start, trace = TRUE)
```

### Models comparison

```{R}
models = list("Exponential model" = mod1, 
              "Logistic model" = mod2, 
              "Gompertz model" = mod3,
              "Richards model" = mod4)
tab = data.frame(loglik = sapply(models, logLik),
                 df = sapply(models, function(m) attr(logLik(m), "df")),
                 Rsquare = sapply(models, function(m) 
                                  cor(data$y, fitted(m))^2),
                 AIC = sapply(models, AIC),
                 AICc = sapply(models, AICc),
                 BIC = sapply(models, BIC))
sel <- apply(tab[,4:6], 2, which.min)
tab$"" <- sapply(tabulate(sel, nbins = length(models))+1, symnum,
                 cutpoints = 0:4, symbols = c("", "*", "**", "***"))
knitr::kable(tab)
```

```{r, echo=FALSE}
# create figure with table for inclusion in README.md
tab <- tableGrob(tab, theme = tab_theme)
ggsave(grid.arrange(tab), 
       file = "COVID-19-IT_infected_table.png", 
       width = 6, height = 1.3, dpi = 300)
```

```{R, out.width="100%"}
ggplot(data, aes(x = date, y = y)) + 
  geom_point() +
  geom_line(aes(y = fitted(mod1), color = "Exponential")) +
  geom_line(aes(y = fitted(mod2), color = "Logistic")) +
  geom_line(aes(y = fitted(mod3), color = "Gompertz")) +
  geom_line(aes(y = fitted(mod4), color = "Richards")) +
  labs(x = "", y = "Infected", color = "Model") +
  scale_color_manual(values = cols) +
  scale_y_continuous(breaks = seq(0, coef(mod2)[1], by = 5000),
                     minor_breaks = seq(0, coef(mod2)[1], by = 1000)) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  theme_bw() +
  theme(legend.position = "top")
```

```{R, out.width="100%"}
last_plot() +
  scale_y_continuous(trans = "log10", limits = c(100,NA)) +
  labs(y = "Infected (log10 scale)")
```

## Predictions

### Point estimates

```{R}
df = data.frame(x = seq(min(data$x), max(data$x)+14))
df = cbind(df, date = as.Date(df$x, origin = data$date[1]-1),
               fit1 = predict(mod1, newdata = df),
               fit2 = predict(mod2, newdata = df),
               fit3 = predict(mod3, newdata = df),
               fit4 = predict(mod4, newdata = df))
ylim = c(0, max(df[,c("fit2", "fit3")]))
```

```{R, echo=-2, out.width="100%"}
ggplot(data, aes(x = date, y = y)) + 
  geom_point() +
  geom_line(data = df, aes(x = date, y = fit1, color = "Exponential")) +
  geom_line(data = df, aes(x = date, y = fit2, color = "Logistic")) +
  geom_line(data = df, aes(x = date, y = fit3, color = "Gompertz")) +
  geom_line(data = df, aes(x = date, y = fit4, color = "Richards")) +
  coord_cartesian(ylim = ylim) +
  labs(x = "", y = "Infected", color = "Model") +
  scale_y_continuous(breaks = seq(0, max(ylim), by = 10000),
                     minor_breaks = seq(0, max(ylim), by = 5000)) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  scale_color_manual(values = cols) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle=60, hjust=1))
```

```{r, eval=FALSE, echo=FALSE}
df = data.frame(x = seq(min(data$x), max(data$x)+90))
df = cbind(df, 
           "y" = c(NA, diff(data$y), rep(NA, nrow(df)-nrow(data))),
           "date" = as.Date(df$x, origin = data$date[1]-1),
           "grad1" = exponentialGrad(mod1, df$x),
           "grad2" = logisticGrad(mod2, df$x),
           "grad3" = gompertzGrad(mod3, df$x))
(peak = c(which(peaks(df$grad2))[1], which(peaks(df$grad3))[1]))

ggplot(df, aes(x = date, y = y)) + 
  geom_point(size = 1) +
  geom_segment(aes(x = date, y = 0, xend = date, yend = y)) +
  geom_line(data = df, aes(x = date, y = grad1, color = "Exponential")) +
  geom_line(data = df, aes(x = date, y = grad2, color = "Logistic")) +
  geom_line(data = df, aes(x = date, y = grad3, color = "Gompertz")) +
  geom_vline(data = df[peak,], aes(xintercept = date), 
             lty = 2, col = cols[2:3]) +
  coord_cartesian(ylim = range(df$y, df$grad2, df$grad3, na.rm = TRUE)) +
  labs(x = "", y = expression(paste(Delta, "Infected")), color = "Model") +
  scale_color_manual(values = cols) +
  annotate(geom = "text", x = df[peak[1],"date"], y = df[peak[1],"grad2"],
           label = df[peak[1],"date"], col = cols[2], 
           hjust = -0.1, vjust = -0.1) +
  annotate(geom = "text", x = df[peak[2],"date"], y = df[peak[2],"grad3"],
           label = df[peak[2],"date"], col = cols[3],
           hjust = -0.1, vjust = -0.1) +
  theme_bw() +
  theme(legend.position = "top")

# calibration based
source("misc/calibrate_nls.R")
calibrate.nls(mod2, coef(mod2)[1]/2, lwr = 0, upr = max(df$x))
calibrate.nls(mod3, coef(mod3)[1]/2, lwr = 0, upr = max(df$x))
# coimpare with 
peak

cal2 = calibrateMBB.nls(mod2, coef(mod2)[1]/2, lwr = 0, upr = max(df$x))
cal2
as.Date(unlist(round(cal2)), origin = data$date[1]-1)

cal3 = calibrateMBB.nls(mod3, coef(mod3)[1]/2, lwr = 0, upr = max(df$x))
cal3
as.Date(unlist(round(cal3)), origin = data$date[1]-1)
```


### Prediction intervals

```{r}
# compute prediction using Moving Block Bootstrap (MBB) for nls
df = data.frame(x = seq(min(data$x), max(data$x)+14))
df = cbind(df, date = as.Date(df$x, origin = data$date[1]-1))

pred1 = cbind(df, "fit" = predict(mod1, newdata = df))
pred1[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod1, df[df$x > max(data$x),])[,2:3]

pred2 = cbind(df, "fit" = predict(mod2, newdata = df))
pred2[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod2, df[df$x > max(data$x),])[,2:3]

pred3 = cbind(df, "fit" = predict(mod3, newdata = df))
pred3[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod3, df[df$x > max(data$x),])[,2:3]

pred4 = cbind(df, "fit" = predict(mod4, newdata = df))
pred4[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod4, df[df$x > max(data$x),])[,2:3]

# predictions for next day
pred = rbind(subset(pred1, x == max(data$x)+1, select = 2:5),
             subset(pred2, x == max(data$x)+1, select = 2:5),
             subset(pred3, x == max(data$x)+1, select = 2:5),
             subset(pred4, x == max(data$x)+1, select = 2:5))
print(pred, digits = 3)

ylim = c(0, max(pred2$upr, pred3$upr, na.rm=TRUE))
```

```{R, echo=-2, out.width="100%"}
ggplot(data, aes(x = date, y = y)) + 
  geom_point() +
  geom_line(data = pred1, aes(x = date, y = fit, color = "Exponential")) +
  geom_line(data = pred2, aes(x = date, y = fit, color = "Logistic")) +
  geom_line(data = pred3, aes(x = date, y = fit, color = "Gompertz")) +
  geom_line(data = pred4, aes(x = date, y = fit, color = "Richards")) +
  geom_ribbon(data = pred1, aes(x = date, ymin = lwr, ymax = upr), 
              inherit.aes = FALSE, fill = cols[1], alpha=0.3) +
  geom_ribbon(data = pred2, aes(x = date, ymin = lwr, ymax = upr), 
              inherit.aes = FALSE, fill = cols[2], alpha=0.3) +
  geom_ribbon(data = pred3, aes(x = date, ymin = lwr, ymax = upr),
              inherit.aes = FALSE, fill = cols[3], alpha=0.3) +
  geom_ribbon(data = pred4, aes(x = date, ymin = lwr, ymax = upr),
              inherit.aes = FALSE, fill = cols[4], alpha=0.3) +
  coord_cartesian(ylim = c(0, max(ylim))) +
  labs(x = "", y = "Infected", color = "Model") +
  scale_y_continuous(minor_breaks = seq(0, max(ylim), by = 10000)) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  scale_color_manual(values = cols) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle=60, hjust=1))
ggsave(file = "COVID-19-IT_infected_plot.png", width = 7, height = 6, dpi = 300)
```

# Modelling total deceased 

```{R}
# create data for analysis
data = data.frame(date = COVID19$data,
                  y = COVID19$deceduti)
data$x = as.numeric(data$date) - min(as.numeric(data$date)) + 1
DT::datatable(data, options = list("pageLength" = 5))
```

## Estimation

### Exponential

```{R}
mod1_start = lm(log(y) ~ x, data = data)
b = unname(coef(mod1_start))
start = list(th1 = exp(b[1]), th2 = b[2])
exponential <- function(x, th1, th2) th1 * exp(th2 * x)
mod1 = nls(y ~ exponential(x, th1, th2), data = data, start = start)
summary(mod1)
```

### Logistic

```{R}
mod2 = nls(y ~ SSlogis(x, Asym, xmid, scal), data = data)
summary(mod2)
```

### Gompertz

```{R}
mod3 = nls(y ~ SSgompertz(x, Asym, b2, b3), data = data)
# manually set starting values
# start = list(Asym = coef(mod2)[1])
# tmp = list(y = log(log(start$Asym) - log(data$y)), x = data$x)
# b = unname(coef(lm(y ~ x, data = tmp)))
# start = c(start, c(b2 = exp(b[1]), b3 = exp(b[2])))
# mod3 = nls(y ~ SSgompertz(x, Asym, b2, b3), data = data, start = start, 
#            control = nls.control(maxiter = 10000))
summary(mod3)
```

### Richards

```{R}
richards <- function(x, th1, th2, th3) th1*(1 - exp(-th2*x))^th3
Loss  <- function(th, y, x) sum((y - richards(x, th[1], th[2], th[3]))^2) 
start <- optim(par = c(coef(mod2)[1], 0.001, 1), fn = Loss, 
               y = data$y, x = data$x)$par
names(start) <- c("th1", "th2", "th3")
mod4 = nls(y ~ richards(x, th1, th2, th3), data = data, start = start,
           # trace = TRUE, algorithm = "port", 
           control = nls.control(maxiter = 1000))
summary(mod4)
```

### Models comparison

```{R}
models = list("Exponential model" = mod1, 
              "Logistic model" = mod2, 
              "Gompertz model" = mod3,
              "Richards model" = mod4)
tab = data.frame(loglik = sapply(models, logLik),
                 df = sapply(models, function(m) attr(logLik(m), "df")),
                 Rsquare = sapply(models, function(m) 
                                  cor(data$y, fitted(m))^2),
                 AIC = sapply(models, AIC),
                 AICc = sapply(models, AICc),
                 BIC = sapply(models, BIC))
sel <- apply(tab[,4:6], 2, which.min)
tab$"" <- sapply(tabulate(sel, nbins = length(models))+1, symnum,
                 cutpoints = 0:4, symbols = c("", "*", "**", "***"))
knitr::kable(tab)
```

```{r, echo=FALSE}
tab <- tableGrob(tab,  theme = tab_theme)
ggsave(grid.arrange(tab), 
       file = "COVID-19-IT_deceased_table.png", 
       width = 6, height = 1.3, dpi = 300)
```


```{R, out.width="100%"}
ggplot(data, aes(x = date, y = y)) + 
  geom_point() +
  geom_line(aes(y = fitted(mod1), color = "Exponential")) +
  geom_line(aes(y = fitted(mod2), color = "Logistic")) +
  geom_line(aes(y = fitted(mod3), color = "Gompertz")) +
  geom_line(aes(y = fitted(mod4), color = "Richards")) +
  labs(x = "", y = "Deceased", color = "Model") +
  scale_color_manual(values = cols) +
  scale_y_continuous(breaks = seq(0, coef(mod2)[1], by = 500),
                     minor_breaks = seq(0, coef(mod2)[1], by = 100)) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  theme_bw() +
  theme(legend.position = "top")
```

```{R, out.width="100%"}
last_plot() +
  scale_y_continuous(trans = "log10", limits = c(10,NA)) +
  labs(y = "Deceased (log10 scale)")
```

## Predictions

### Point estimates

```{R}
df = data.frame(x = seq(min(data$x), max(data$x)+14))
df = cbind(df, date = as.Date(df$x, origin = data$date[1]-1),
               fit1 = predict(mod1, newdata = df),
               fit2 = predict(mod2, newdata = df),
               fit3 = predict(mod3, newdata = df),
               fit4 = predict(mod4, newdata = df))
ylim = c(0, max(df[,-(1:3)]))
```

```{R, out.width="100%"}
ggplot(data, aes(x = date, y = y)) + 
  geom_point() +
  geom_line(data = df, aes(x = date, y = fit1, color = "Exponential")) +
  geom_line(data = df, aes(x = date, y = fit2, color = "Logistic")) +
  geom_line(data = df, aes(x = date, y = fit3, color = "Gompertz")) +
  geom_line(data = df, aes(x = date, y = fit4, color = "Richards")) +
  coord_cartesian(ylim = ylim) +
  labs(x = "", y = "Deceased", color = "Model") +
  scale_y_continuous(breaks = seq(0, max(ylim), by = 1000),
                     minor_breaks = seq(0, max(ylim), by = 1000)) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  scale_color_manual(values = cols) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle=60, hjust=1))
```

```{r, eval=FALSE, echo=FALSE}
df = data.frame(x = seq(min(data$x), max(data$x)+90))
df = cbind(df, 
           "y" = c(NA, diff(data$y), rep(NA, nrow(df)-nrow(data))),
           "date" = as.Date(df$x, origin = data$date[1]-1),
           "grad1" = exponentialGrad(mod1, df$x),
           "grad2" = logisticGrad(mod2, df$x),
           "grad3" = gompertzGrad(mod3, df$x))
           # "grad4" = richardsGrad(mod4, df$x))
peak = c(which(peaks(df$grad2)), which(peaks(df$grad3)))

ggplot(df, aes(x = date, y = y)) + 
  geom_point() +
  geom_segment(aes(x = date, y = 0, xend = date, yend = y)) +
  geom_line(data = df, aes(x = date, y = grad1, color = "Exponential")) +
  geom_line(data = df, aes(x = date, y = grad2, color = "Logistic")) +
  geom_line(data = df, aes(x = date, y = grad3, color = "Gompertz")) +
  # geom_line(data = df, aes(x = date, y = grad4, color = "Richards")) +
  geom_vline(data = df[peak,], aes(xintercept = date), lty = 2) +
  coord_cartesian(ylim = range(df$grad2, df$grad3, df$grad4)) +
  labs(x = "", y = expression(paste(Delta, "Deceased")), color = "Model") +
  scale_color_manual(values = cols) +
  annotate(geom = "text", x = df[peak[1],"date"], y = df[peak[1],"grad2"],
           label = df[peak[1],"date"], col = cols[2], 
           hjust = -0.1, vjust = -0.1) +
  annotate(geom = "text", x = df[peak[2],"date"], y = df[peak[2],"grad3"],
           label = df[peak[2],"date"], col = cols[3],
           hjust = -0.1, vjust = -0.1) +
  # annotate(geom = "text", x = df[peak[3],"date"], y = df[peak[3],"grad4"],
  #          label = df[peak[3],"date"], col = cols[4],
  #          hjust = -0.1, vjust = -0.1) +
  theme_bw() +
  theme(legend.position = "top")
```


### Prediction intervals

```{r}
# compute prediction using Moving Block Bootstrap (MBB) for nls
df = data.frame(x = seq(min(data$x), max(data$x)+14))
df = cbind(df, date = as.Date(df$x, origin = data$date[1]-1))

pred1 = cbind(df, "fit" = predict(mod1, newdata = df))
pred1[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod1, df[df$x > max(data$x),])[,2:3]

pred2 = cbind(df, "fit" = predict(mod2, newdata = df))
pred2[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod2, df[df$x > max(data$x),])[,2:3]

pred3 = cbind(df, "fit" = predict(mod3, newdata = df))
pred3[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod3, df[df$x > max(data$x),])[,2:3]

pred4 = cbind(df, "fit" = predict(mod4, newdata = df))
pred4[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod4, df[df$x > max(data$x),])[,2:3]

# predictions for next day
pred = rbind(subset(pred1, x == max(data$x)+1, select = 2:5),
             subset(pred2, x == max(data$x)+1, select = 2:5),
             subset(pred3, x == max(data$x)+1, select = 2:5),
             subset(pred4, x == max(data$x)+1, select = 2:5))
print(pred, digits = 3)

ylim = c(0, max(pred2$upr, pred3$upr, na.rm=TRUE))
```

```{R, echo=-2, out.width="100%"}
ggplot(data, aes(x = date, y = y)) + 
  geom_point() +
  geom_line(data = pred1, aes(x = date, y = fit, color = "Exponential")) +
  geom_line(data = pred2, aes(x = date, y = fit, color = "Logistic")) +
  geom_line(data = pred3, aes(x = date, y = fit, color = "Gompertz")) +
  geom_line(data = pred4, aes(x = date, y = fit, color = "Richards")) +
  geom_ribbon(data = pred1, aes(x = date, ymin = lwr, ymax = upr), 
              inherit.aes = FALSE, fill = cols[1], alpha=0.3) +
  geom_ribbon(data = pred2, aes(x = date, ymin = lwr, ymax = upr), 
              inherit.aes = FALSE, fill = cols[2], alpha=0.3) +
  geom_ribbon(data = pred3, aes(x = date, ymin = lwr, ymax = upr),
              inherit.aes = FALSE, fill = cols[3], alpha=0.3) +
  geom_ribbon(data = pred4, aes(x = date, ymin = lwr, ymax = upr),
              inherit.aes = FALSE, fill = cols[4], alpha=0.3) +
  coord_cartesian(ylim = c(0, max(ylim))) +
  labs(x = "", y = "Deceased", color = "Model") +
  scale_y_continuous(minor_breaks = seq(0, max(ylim), by = 1000)) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  scale_color_manual(values = cols) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle=60, hjust=1))
ggsave(file = "COVID-19-IT_deceased_plot.png", width = 7, height = 6, dpi = 300)
```

# Modelling recovered

```{R}
# create data for analysis
data = data.frame(date = COVID19$data,
                  y = COVID19$dimessi_guariti)
data$x = as.numeric(data$date) - min(as.numeric(data$date)) + 1
DT::datatable(data, options = list("pageLength" = 5))
```

## Estimation

### Exponential

```{R}
mod1_start = lm(log(y) ~ x, data = data)
b = unname(coef(mod1_start))
start = list(th1 = exp(b[1]), th2 = b[2])
exponential <- function(x, th1, th2) th1 * exp(th2 * x)
mod1 = nls(y ~ exponential(x, th1, th2), data = data, start = start)
summary(mod1)
```

### Logistic

```{R}
mod2 = nls(y ~ SSlogis(x, Asym, xmid, scal), data = data)
summary(mod2)
```

### Gompertz

```{R}
mod3 = nls(y ~ SSgompertz(x, Asym, b2, b3), data = data)
summary(mod3)
```

### Richards

```{R}
richards <- function(x, th1, th2, th3) th1*(1 - exp(-th2*x))^th3
Loss  <- function(th, y, x) sum((y - richards(x, th[1], th[2], th[3]))^2) 
start <- optim(par = c(coef(mod2)[1], 0.001, 1), fn = Loss, 
               y = data$y, x = data$x)$par
names(start) <- c("th1", "th2", "th3")
mod4 = nls(y ~ richards(x, th1, th2, th3), data = data, start = start,
           # trace = TRUE, # algorithm = "port", 
           control = nls.control(maxiter = 1000))
summary(mod4)
```

### Models comparison

```{R}
models = list("Exponential model" = mod1, 
              "Logistic model" = mod2, 
              "Gompertz model" = mod3,
              "Richards model" = mod4)
tab = data.frame(loglik = sapply(models, logLik),
                 df = sapply(models, function(m) attr(logLik(m), "df")),
                 Rsquare = sapply(models, function(m) 
                                  cor(data$y, fitted(m))^2),
                 AIC = sapply(models, AIC),
                 AICc = sapply(models, AICc),
                 BIC = sapply(models, BIC))
sel <- apply(tab[,4:6], 2, which.min)
tab$"" <- sapply(tabulate(sel, nbins = length(models))+1, symnum,
                 cutpoints = 0:4, symbols = c("", "*", "**", "***"))
knitr::kable(tab)
```

```{r, echo=FALSE}
tab <- tableGrob(tab,  theme = tab_theme)
ggsave(grid.arrange(tab), 
       file = "COVID-19-IT_recovered_table.png", 
       width = 6, height = 1.3, dpi = 300)
```

```{R, out.width="100%"}
ggplot(data, aes(x = date, y = y)) + 
  geom_point() +
  geom_line(aes(y = fitted(mod1), color = "Exponential")) +
  geom_line(aes(y = fitted(mod2), color = "Logistic")) +
  geom_line(aes(y = fitted(mod3), color = "Gompertz")) +
  geom_line(aes(y = fitted(mod4), color = "Richards")) +
  labs(x = "", y = "Recovered", color = "Model") +
  scale_color_manual(values = cols) +
  scale_y_continuous(breaks = seq(0, coef(mod2)[1], by = 500),
                     minor_breaks = seq(0, coef(mod2)[1], by = 100)) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  theme_bw() +
  theme(legend.position = "top")
```

```{R, out.width="100%"}
last_plot() +
  scale_y_continuous(trans = "log10", limits = c(10,NA)) +
  labs(y = "Recovered (log10 scale)")
```

## Predictions

### Point estimates

```{R}
df = data.frame(x = seq(min(data$x), max(data$x)+14))
df = cbind(df, date = as.Date(df$x, origin = data$date[1]-1),
               fit1 = predict(mod1, newdata = df),
               fit2 = predict(mod2, newdata = df),
               fit3 = predict(mod3, newdata = df),
               fit4 = predict(mod4, newdata = df))
ylim = c(0, max(df[,-(1:3)]))
```

```{R, out.width="100%"}
ggplot(data, aes(x = date, y = y)) + 
  geom_point() + 
  geom_line(data = df, aes(x = date, y = fit1, color = "Exponential")) +
  geom_line(data = df, aes(x = date, y = fit2, color = "Logistic")) +
  geom_line(data = df, aes(x = date, y = fit3, color = "Gompertz")) +
  geom_line(data = df, aes(x = date, y = fit4, color = "Richards")) +
  coord_cartesian(ylim = ylim) +
  labs(x = "", y = "Recovered", color = "Model") +
  scale_y_continuous(breaks = seq(0, max(ylim), by = 1000),
                     minor_breaks = seq(0, max(ylim), by = 1000)) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  scale_color_manual(values = cols) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle=60, hjust=1))
```

```{r, echo=FALSE, eval=FALSE}
df = data.frame(x = seq(min(data$x), max(data$x)+90))
df = cbind(df, 
           "y" = c(NA, diff(data$y), rep(NA, nrow(df)-nrow(data))),
           "date" = as.Date(df$x, origin = data$date[1]-1),
           "grad1" = exponentialGrad(mod1, df$x),
           "grad2" = logisticGrad(mod2, df$x),
           "grad3" = gompertzGrad(mod3, df$x))
peak = c(which(peaks(df$grad2)), which(peaks(df$grad3)))

ggplot(df, aes(x = date, y = y)) + 
  geom_point() +
  geom_segment(aes(x = date, y = 0, xend = date, yend = y)) +
  geom_line(data = df, aes(x = date, y = grad1, color = "Exponential")) +
  geom_line(data = df, aes(x = date, y = grad2, color = "Logistic")) +
  geom_line(data = df, aes(x = date, y = grad3, color = "Gompertz")) +
  geom_vline(data = df[peak,], aes(xintercept = date), 
             lty = 2, col = cols[2]) +
  coord_cartesian(ylim = c(0,range(df$grad2[peak],df$grad3[peak]))) +
  labs(x = "", y = expression(paste(Delta, "Recovered")), color = "Model") +
  scale_color_manual(values = cols) +
  annotate(geom = "text", x = df[peak[1],"date"], y = df[peak[1],"grad2"],
           label = df[peak[1],"date"], col = cols[2], 
           hjust = -0.1, vjust = -0.1) +
  annotate(geom = "text", x = df[peak[2],"date"], y = df[peak[2],"grad3"],
           label = df[peak[2],"date"], col = cols[3],
           hjust = -0.1, vjust = -0.1) +
  theme_bw() +
  theme(legend.position = "top")
```

### Prediction intervals

```{r}
# compute prediction using Moving Block Bootstrap (MBB) for nls
df = data.frame(x = seq(min(data$x), max(data$x)+14))
df = cbind(df, date = as.Date(df$x, origin = data$date[1]-1))

pred1 = cbind(df, "fit" = predict(mod1, newdata = df))
pred1[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod1, df[df$x > max(data$x),])[,2:3]

pred2 = cbind(df, "fit" = predict(mod2, newdata = df))
pred2[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod2, df[df$x > max(data$x),])[,2:3]

pred3 = cbind(df, "fit" = predict(mod3, newdata = df))
pred3[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod3, df[df$x > max(data$x),])[,2:3]

pred4 = cbind(df, "fit" = predict(mod4, newdata = df))
pred4[df$x > max(data$x), c("lwr", "upr")] = predictMBB.nls(mod4, df[df$x > max(data$x),])[,2:3]

# predictions for next day
pred = rbind(subset(pred1, x == max(data$x)+1, select = 2:5),
             subset(pred2, x == max(data$x)+1, select = 2:5),
             subset(pred3, x == max(data$x)+1, select = 2:5),
             subset(pred4, x == max(data$x)+1, select = 2:5))
print(pred, digits = 3)

ylim = c(0, max(pred2$upr, pred3$upr, na.rm=TRUE))
```

```{R, echo=-2, out.width="100%"}
ggplot(data, aes(x = date, y = y)) + 
  geom_point() +
  geom_line(data = pred1, aes(x = date, y = fit, color = "Exponential")) +
  geom_line(data = pred2, aes(x = date, y = fit, color = "Logistic")) +
  geom_line(data = pred3, aes(x = date, y = fit, color = "Gompertz")) +
  geom_line(data = pred4, aes(x = date, y = fit, color = "Richards")) +
  geom_ribbon(data = pred1, aes(x = date, ymin = lwr, ymax = upr), 
              inherit.aes = FALSE, fill = cols[1], alpha=0.3) +
  geom_ribbon(data = pred2, aes(x = date, ymin = lwr, ymax = upr), 
              inherit.aes = FALSE, fill = cols[2], alpha=0.3) +
  geom_ribbon(data = pred3, aes(x = date, ymin = lwr, ymax = upr),
              inherit.aes = FALSE, fill = cols[3], alpha=0.3) +
  geom_ribbon(data = pred4, aes(x = date, ymin = lwr, ymax = upr),
              inherit.aes = FALSE, fill = cols[4], alpha=0.3) +
  coord_cartesian(ylim = c(0, max(ylim))) +
  labs(x = "", y = "Recovered", color = "Model") +
  scale_y_continuous(breaks = seq(0, max(ylim), by = 5000),
                     minor_breaks = seq(0, max(ylim), by = 1000)) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  scale_color_manual(values = cols) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle=60, hjust=1))
ggsave(file = "COVID-19-IT_recovered_plot.png", width = 7, height = 6, dpi = 300)
```

# Evolution of positive cases and administered swabs

```{R}
df = data.frame(date = COVID19$data,
                positives = c(NA, diff(COVID19$totale_casi)),
                swabs = c(NA, diff(COVID19$tamponi)))
df$x = as.numeric(df$date) - min(as.numeric(df$date)) + 1
df$y = df$positives/df$swabs
df = subset(df, swabs > 50)
DT::datatable(df[,-4], )
```

<br>

```{R}
ggplot(df, aes(x = date)) + 
  geom_point(aes(y = swabs, color = "swabs"), pch = 19) +
  geom_line(aes(y = swabs, color = "swabs")) +
  geom_point(aes(y = positives, color = "positives"), pch = 0) +
  geom_line(aes(y = positives, color = "positives")) +
  labs(x = "", y = "Number of cases", color = "") +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  scale_color_manual(values = palette()[c(2,1)]) +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle=60, hjust=1))
```

```{R}
ggplot(df, aes(x = date, y = y)) + 
  geom_smooth(method = "loess", se = TRUE, col = "darkgrey") +
  geom_point(col=palette()[4]) + 
  geom_line(size = 0.5, col=palette()[4]) +
  labs(x = "", y = "Positives / Admnistered swabs") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_date(date_breaks = "2 day", date_labels =  "%b%d",
               minor_breaks = "1 day") +
  theme_bw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle=60, hjust=1))
```



